Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review

### 2017-12-04 analysis/paper/paper.Rmd
references update, Glucagon WIP

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: mee.csl
abstract: |
  (maximum 200 words).

@@

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(refractoryANA)
require(magrittr,quietly = T)

@@

# Introduction

Anaphylaxis is a life-threatening immunologic reaction to an allergen (most frequently food, drugs, insect venom) that is hardly avoidable in clinical practice of various medical specialties. Anaphylaxis is considered a clinical diagnosis and it is highly likely if there is an acute onset of skin symptoms with either hypotension or respiratory compromise, or when the allergen is known or likely [@Sampson2006]. Its pathogenesis relies on a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

Anaphylaxis remains a clinical diagnosis [@Kemp2014], but consensus criteria of anaphylaxis suggest a standardized set of criteria [@Sampson2006]. The definition of _refractory_ anaphylaxis is not established. For the purpose of this review, we will define it as anaphylaxis (meeting the criteria by NIAID/FAAN [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

It is important to distinguish refractory anaphylaxis from recurrent anaphylaxis. In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab. We did not include such reports in our analysis and advise on using the term "recurrent" anaphylaxis instead of "refractory" in such cases.

@@

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is usually food dependent and less severe than episodes in adulthood. Cases of _refractory_ anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our experience) to be 3-5% of all cases of anaphylaxis.

@@

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. The mainstay of therapy is the early administration of intramuscular (i.m.) adrenalin [@Lazar2017].
However, optimal anaphylaxis treatment is difficult to study because it is an emergency condition and placebo-controlled studies are impossible. Therefore most of the knowledge is based on expert recommendations which are graded as low-level evidence sources.

We aimed to reevaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenalin were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenalin. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included 29 Papers reporting ` r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenalin. Study outcomes were defined as time to elicit a response and mortality. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports therefore the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration.

@@

```{r demograph}
source("../../R/data_handling.R")
knitr::kable(demogr.tab,col.names = c("Elicitor","n","Age","Male fraction","Operative fraction"),digits = 1)
```

@@

The risk factors for refractory anaphylaxis include perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the reaction happened during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Surprisingly, no cases of insect venom anaphylaxis were found among these reports.cWhile the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities. [@Nag2017]

<!-- # Firs sings in anesthesia setting of anaphylaxis -->
Early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.

@@

World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011]. Patients should be transported to a critical care unit. AAAAI [@Lieberman] suggest similar treatment but include dopamine, vasopressin if epinephrine injections and volume expansion fail to alleviate hypotension, glucagon atropine, and methylene blue. EAACI [@Muraro2014] guidelines also suggest using glucagon in case of concomitant beta-blockade.

@@

Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013]. Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) *[7-9]* production which is a potent mediator of anaphylaxis [].

@@

As methylene blue does not block nitric oxide production its side effects are minimal (Nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis. No side effects were reported due to methylene blue use in the above cases.

It is important to remember that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].

# Vasopressors Vasopressin as a supplementary vasopressor

@@

Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol, methoxamine, dopamine and noradrenaline infusions.

Our register lists 18 cases where dopamine was used. None were fatal.

Use of vasopressin in refractory anaphylaxis was only suggested in the AAAAI management criteria [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine[@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

Seedat et al. described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia. The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone. Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred. Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued. [@Seedat2014] We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.

@@

@@

Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986, @Javeed1996].  Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986], Glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.

@@

## What are the key weaknesses in clinical management so far?

## What potential does further research hold? What is the ultimate goal in this field?

## What research or knowledge is needed to achieve this goal and what is the biggest challenge in this goal being achieved?

## Is there any particular area of the research you are finding of interest at present?
<!-- **Please note that ‘opinions’ are encouraged in the Expert commentary section, and, as such, referees are asked to keep this in mind when peer reviewing the manuscript.** -->

# Five-year view
<!-- **Authors are challenged to include a speculative viewpoint on how the field will have evolved five years from the point at which the review was written.** -->

@@

```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
source("../../R/prisma.R")
knitr::include_graphics("../figures/prism_flow.png")
```

>>>>

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: mee.csl
abstract: |
  (maximum 200 words).

@@

```{r, setup, echo = FALSE,include=F}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

#library(refractoryANA)
require(magrittr,quietly = T)
# source("../../R/data_handling.R")
devtools::load_all(".",quiet = T)

@@

# Introduction
Anaphylaxis is a potentially life-threatening mast cell dependent reaction to  most frequently food, drugs, insect venom. Anaphylaxis is a clinical diagnosis. It is highly likely if there is an acute onset of skin symptoms in conjuction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].By contrast the term refractory anaphylaxis has not established. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

It is important to distinguish refractory anaphylaxis from recurrent anaphylaxis. In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab. We did not include such reports into our analysis and suggest to use the term “recurrent” anaphylaxis instead of “refractory” in such cases.

@@

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is in the majority of cases food dependent and most likely more frequently less severe than episodes in adulthood (Ref). Cases of refractory anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3-5% of all cases of anaphylaxis.

@@

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. Until to day the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are for ethical reasons limited. Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.

In the present analysis we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included 29 Papers reporting `r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenaline. Study outcomes were defined as time to elicit a response and mortality. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports therefore the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration.

@@

```{r demograph}
knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1)
```

@@

A major risk factor for refractory anaphylaxis is perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occured during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Surprisingly, no cases of insect venom anaphylaxis were found among these reports. While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 – 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 – 10% of the perioperative fatalities. [@Nag2017]

<!-- # Firs sings in anesthesia setting of anaphylaxis -->
An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.

@@

World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011]. Patients should be transported to a critical care unit. The AAAAI [@Lieberman] is  suggesting a similar treatment but include dopamine and vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension. Also the EAACI [@Muraro2014] guidelines suggest to use glucagon in case of concomitant beta-blockade.

@@

Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013]. Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) production which is a potent mediator of anaphylaxis [@Moritoki1992].

@@

As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis. No side effects were reported due to methylene blue use in the above cases.

Clinicians must be aware that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].

# Vasopressors

@@

Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol 
[`r which.sources(d$Source,"metara",list(d$What.helped,d$Therapy))`], methoxamine 
[`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
Drugs given perioperatively were the elicing factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug. One patient rected to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].

Use of vasopressin in refractory anaphylaxis was only suggested in  management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

Seedat et al. described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia. The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone [@Seedat2014]. Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred. Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014]. We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.

@@

The European anaphylaxis register lists 18 cases where dopamine was used. None of these reactions were fatal and  Ring and Messmer. `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure. Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%

@@

Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Patients

Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986], Glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.

@@

Severe anaphylaxis is a critical medical condition requiring an immediate intervention. According to the international guidelines adrenaline intramuscular given is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient. However, in a few patients this intervention might not be sufficient and besides repeating the doses other therapeutical measures will need to be applied. Such cases are regarded as refractory anaphylaxis, which are particulary seen, when the elicitor of the reaction is reaching the organism intravenously. This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the ….clotting …system?? check. 
In clinical practice it is necessary to identify patients of risk. Such risk factors should particulary be assessed in patients undergoing medical diagnostic and therapeutic measures. The measurement of tryptase enables to identify patients how are at risk for severe anaphylaxis.

For the future, biomarker measurements in a bed side setting may enable doctors to identify  patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment. Moreover, defined standardized treatment protocols for such cases of refractory anaphylaxis may provide an optimization of the treatment which can safe life in a given patient.

@@

```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
#source("../../R/prisma.R")
knitr::include_graphics("../figures/prism_flow.png")
```

```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
#source("../../R/prisma.R")
knitr::include_graphics("../figures/algo.png")
```

```{r dopamine}
knitr::kable(dopam.anareg[,c(-3,-5)])
```
